Better fusion for filters

### 2014-04-02 2014betterfusionforfilters/section/04-ILP.tex
objective function
explain ILP variables

\begin{tabbing}
MMMM        \= MM   \= \kill
$nodes$     \> @::@ \> $program \to V$          \\
$edges$     \> @::@ \> $program \to E$          \\
$edge$      \> @::@ \> $\{bind\} \times bind \to E$\\
\\
MMMMM       \= \kill
$nodes(bs)$ \> $= \{(name(b), \tau_b) | b \in bs\}$       \\
\\
$edges(bs)$ \> $= \bigcup_{b \in bs}edge(b)$    \\
\\
MMMM             \= M \= \kill
$edge(bs, out = @fold@~f~in)$ \\
                              \> $=$    \> $\{inedge(bs,out,s) | s \in f\}$ \\
                              \> $\cup$ \> $\{inedge(bs, out, in) \}$       \\
$edge(bs, out = @map@~f~in)$  \\
                              \> $=$    \> $\{inedge(bs,out,s) | s \in f\}$ \\
                              \> $\cup$ \> $\{inedge(bs, out, in) \}$       \\
$edge(bs, out = @filter@~f~in)$  \\
                              \> $=$    \> $\{inedge(bs,out,s) | s \in f\}$ \\
                              \> $\cup$ \> $\{inedge(bs, out, in) \}$       \\
$edge(bs, out = @gather@~data~indices)$  \\
                              \> $=$    \> $\{(out,data, fusion~preventing) \}$ \\
                              \> $\cup$ \> $\{inedge(bs, out, indices) \}$       \\
$edge(bs, out = @cross@~a~b)$            \\
                              \> $=$    \> $\{inedge(bs, out, a) \}$       \\
                              \> $\cup$ \> $\{(out, b, fusion~preventing) \}$ \\
$edge(bs, outs = @external@~ins)$  \\
                              \> $=$    \> $\{(outs,i, fusion~preventing) | i \in ins \}$ \\
\\
$inedge(bs,to,from)$ \\
                     \> $|$ \> $(from = @fold@~f~s) \in bs$     \\
                     \> $=$ \> $(to, from, fusion~preventing)$  \\
                     \> $|$ \> $(\ldots,from,\ldots = @external@ \ldots) \in bs$     \\
                     \> $=$ \> $(to, from, fusion~preventing)$  \\
                     \> $|$ \> $otherwise$                      \\
                     \> $=$ \> $(to, from, fusible)$            \\
\end{tabbing}

\subsection{Graph $\to$ ILP}

ILP variables:
\begin{tabbing}
M   \= M \= MMMMMMM \= MM \= \kill
$x$   \> @::@  \> $node \times node$ \> $\to$ \> $\mathbb{B}$ \\
      \>       \> ($x_{ij} = 0$ iff $i$ and $j$ are fused together) \\
$\pi$ \> @::@  \> $node$             \> $\to$ \> $\mathbb{R}$ \\
      \>       \> ($\pi_i$ is used to ensure an acyclic clustering) \\
$c$   \> @::@  \> $node$             \> $\to$ \> $\mathbb{B}$ \\
      \>       \> ($c_i = 0$ iff $\forall (i,j) \in E. x_{ij} = 0$) \\
      \>       \> (i.e. no intermediate array necessary)            \\
\end{tabbing}

Talk about how $x_{ij} = 0$ if and only if $i$ and $j$ are fused together, otherwise $x_{ij} = 1$. This may seem reversed, but makes certain constraints easier to formulate.

Use the lemma:
a graph $G$ is acyclic if and only if there exists some mapping $\pi :: node \to \mathbb{R}$ such that for every edge in $G$ from $i$ to $j$, $\pi_i < \pi_j$.

@@

\subsubsection{The benefits of fusion}

Fusing loops can have three main benefits:
\begin{itemize}
\item
reducing memory traffic, such as multiple loops reading from the same array;
\item
removing intermediate arrays, thus reducing the amount of memory required;
\item
and reducing loop overhead, such as when two loops operate on different arrays of the same size.
\end{itemize}
However, these benefits cannot be considered in isolation; for example, fusing two loops to reduce loop overhead may remove potential fusion opportunities that reduce memory traffic.
When operating on large arrays that do not fit in cache, memory traffic dominates execution time.
An excessive number of intermediate arrays can also cause issues if all are live in memory at once, potentially leading to \emph{thrashing}.
The benefits of removing loop overhead are least of all; it should be performed if possible, but must never remove opportunities for other fusion.

>>>>

\begin{tabbing}
MMMM        \= MM   \= \kill
$nodes$     \> @::@ \> $program \to V$          \\
$edges$     \> @::@ \> $program \to E$          \\
$edge$      \> @::@ \> $\{bind\} \times bind \to E$\\
\\
MMMMM       \= \kill
$nodes(bs)$ \> $= \{(name(b), \tau_b) | b \in bs\}$       \\
\end{tabbing}
Each node may simply be the name of its output binding (or bindings, in the case of @external@); as we require names to only be bound once, this is assured to be unique.
Creating edges between these nodes is simply when one binding references an earlier one. The only complication is designating edges as \emph{fusible} or \emph{fusion-preventing}.

\begin{tabbing}
MMMMM       \= \kill
$edges(bs)$ \> $= \bigcup_{b \in bs}edge(b)$    \\
\\
MMMM             \= M \= \kill
$edge(bs, out = @fold@~f~in)$ \\
                              \> $=$    \> $\{inedge(bs,out,s) | s \in f\}$ \\
                              \> $\cup$ \> $\{inedge(bs, out, in) \}$       \\
$edge(bs, out = @map@~f~in)$  \\
                              \> $=$    \> $\{inedge(bs,out,s) | s \in f\}$ \\
                              \> $\cup$ \> $\{inedge(bs, out, in) \}$       \\
$edge(bs, out = @filter@~f~in)$  \\
                              \> $=$    \> $\{inedge(bs,out,s) | s \in f\}$ \\
                              \> $\cup$ \> $\{inedge(bs, out, in) \}$       \\
$edge(bs, out = @gather@~data~indices)$  \\
                              \> $=$    \> $\{(out,data, fusion~preventing) \}$ \\
                              \> $\cup$ \> $\{inedge(bs, out, indices) \}$       \\
$edge(bs, out = @cross@~a~b)$            \\
                              \> $=$    \> $\{inedge(bs, out, a) \}$       \\
                              \> $\cup$ \> $\{(out, b, fusion~preventing) \}$ \\
$edge(bs, outs = @external@~ins)$  \\
                              \> $=$    \> $\{(outs,i, fusion~preventing) | i \in ins \}$ \\
\\
$inedge(bs,to,from)$ \\
                     \> $|$ \> $(from = @fold@~f~s) \in bs$     \\
                     \> $=$ \> $(to, from, fusion~preventing)$  \\
                     \> $|$ \> $(\ldots,from,\ldots = @external@ \ldots) \in bs$     \\
                     \> $=$ \> $(to, from, fusion~preventing)$  \\
                     \> $|$ \> $otherwise$                      \\
                     \> $=$ \> $(to, from, fusible)$            \\
\end{tabbing}

\subsection{Graph $\to$ ILP}
With the dependency graph now constructed, it must be converted to an integer linear program.
The integer linear program will contain some variables, an objective function, and constraints that the variables must conform to.
An external solver will find and return a variable assignment that minimises the objective function.
With the variable assignment, the graph's nodes are partitioned into a clustering, and then flow fusion\cite{lippmeier2013flow} is used to extract imperative loops for each cluster.

The variables in the ILP formulation can be split into three groups:

\begin{tabbing}
M   \= MM \= MMMMMMM \= MM \= \kill
$x$   \> @::@  \> $node \times node$ \> $\to$ \> $\mathbb{B}$
\end{tabbing}
The first and most important group is created for every pair of nodes, and has a boolean value indicating whether the two nodes can be fused.
If $x_{ij} = 0$, then $i$ and $j$ are fused together.
While it may seem counterintuitive for 0 to mean fused and 1 unfused, it means the objective function can simply minimise a weighted $x_{ij}$, and also makes later precedence constraints simpler.
The values of these variables are used to construct the partitioning at the end, such that $\forall i,j.\ x_{ij} = 0 \iff cluster(i) = cluster(j)$.

\begin{tabbing}
M   \= MM \= MMMMMMM \= MM \= \kill
$\pi$ \> @::@  \> $node$             \> $\to$ \> $\mathbb{R}$
\end{tabbing}
The second group of variables is used simply to ensure that the resulting clustering is acyclic.
For some node $i$, we associate a real $\pi_i$ such that every node $j$ after $i$ has a value $\pi_j > \pi_i$.
This has the effect of requiring an acyclic clustering, as if one node is after another its $\pi$ must be strictly greater than its predecessor, and the predecessor must be strictly less than the successor.
If two nodes are to be fused together, their $\pi$ values must be the same.

\begin{tabbing}
M   \= MM \= MMMMMMM \= MM \= \kill
$c$   \> @::@  \> $node$             \> $\to$ \> $\mathbb{B}$
\end{tabbing}
The third, and final, group of variables are purely for minimisation, indicating whether a node's array is \emph{fully contracted}.
They are not required for correctness of the clustering, but are used in the objective function.
A node's array will be fully contracted --- that is, it will be replaced by a scalar --- if all nodes that use this array are fused together:
$c_i = 0 \iff \forall (i,j) \in E. x_{ij} = 0$.

@@

\subsubsection{The objective function}

To find the objective function, note that fusing loops can have three main benefits:
\begin{itemize}
\item
reducing memory traffic, such as multiple loops reading from the same array;
\item
removing intermediate arrays, thus reducing the amount of memory required;
\item
and reducing loop overhead, such as when two loops operate on different arrays of the same size.
\end{itemize}
However, these benefits cannot be considered in isolation; for example, fusing two loops to reduce loop overhead may remove potential fusion opportunities that reduce memory traffic.
When operating on large arrays that do not fit in cache, memory traffic dominates execution time.
An excessive number of intermediate arrays can also cause issues if all are live in memory at once, potentially leading to \emph{thrashing}.
The benefits of removing loop overhead are least of all; it should be performed if possible, but must never remove opportunities for other fusion.
